---
title: "Week 4 Project"
author: "Nathan Young"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(caret)
library(randomForest)
library(gt)
knitr::opts_chunk$set(echo = TRUE)
```

## Background  
Using devices such as *Jawbone Up*, *Nike FuelBand*, and *Fitbit* it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 (classe variable) different ways. More information is available from the [website](http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har) 

 (see the section on the Weight Lifting Exercise Dataset).

## Data 
Data for this project was retrieved from the proved url's for training and test data. To prevent accessing the data repeatedly, it's good practice to surround a download call with a test if the file already exists. When loading the data, a warning was raised that there were parsing issues; when `problems(data)` was called, I found that there were several `#DIV/0!` strings that were preventing a clean import. This was managed by including the string in the na character list. 
```{r, cache = TRUE}
url_train <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
url_test <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

if (!file.exists("./data/train.csv")) {
  download.file(url_train, destfile = "./data/train.csv")
}

if (!file.exists("./data/test.csv")) {
  download.file(url_test, destfile = "./data/test.csv")
}

train <- read_csv("./data/train.csv", na = c("NA", "", "#DIV/0!"))
test <- read_csv("./data/test.csv", na = c("NA", "", "#DIV/0!"))
```
## Exploration 
The first step in this process was to identify which variables in the dataset are useful for the analysis. When looking at a summary of the dataset, many of the columns were mostly `NA` so these columns were removed. Additionally, the predictor "classe", and columns 1 - 7 can be removed as these contain information not relevant to predicting "classe". Note that for the analyses without PCA, the `classe` variable is left in the train set. 

```{r data cleaning, echo = FALSE}
# First subset out classe and 1-7 and classe for PCA
train_subset <- train[,-c(1:7)] %>%
  # Remove columns that consist of mostly NA values
  discard(~(sum(is.na(.))/nrow(train) > .8))

# positive correlation matrix
M <- abs(cor(train_subset[,-53]))
# Set diagonal to zero as we don't care that they are 1
diag(M) <- 0
ncorr <- nrow(which(M > 0.8, arr.ind = T))
```

## Analysis  
This is a classification problem where the goal is to classify the data into one of 5 groups: A, B, C, D, or E. As such, a decision tree from `randomForest` is a classic choice. There are also `r ncorr` variables that are highly correlated with each other so doing a preprocess with PCA would reduce dimension of dataset. To evaluate which approach is best, I decided to build three models: 

1. A classification tree with `caret::train` and `rpart` using a PCA preprocess step. 
2. A classification tree identical to 1. but without PCA
3. A random forest model with `randomForest::randomForest`

Cross validation is included in the first two models by passing an argument to `trControl = trainControl(method = "cv", number = 3)`. 

To measure out of sample error rates, the provided training data was split into a further training and validation set. The out of sample error rate is calculated as the `1 - Accuracy` of each model on the validation set. 

```{r validation}
# Split training into train and validation sets. 
inTrain <- createDataPartition(y=train_subset$classe, p=0.7, list=F)
train_subset_2 <- train_subset[inTrain,]
valid <- train_subset[-inTrain,]
```

```{r PCA Model}
# PCA without "classe"
preProc <- preProcess(train_subset_2[,-53],method="pca")
trainPC <- cbind(classe = train_subset_2$classe, 
                 predict(preProc,train_subset_2[,-53]))
validPC <- cbind(classe = valid$classe, 
                 predict(preProc, valid[,-53]))

mdl_rpart_PC <- train(classe ~ .,
                      method="rpart",
                      trControl = trainControl(method = "cv", number = 3),
                      data=trainPC)
pred_rpart_PC <- predict(mdl_rpart_PC, trainPC)
cM_rpart_PC <- confusionMatrix(as.factor(train_subset_2$classe), pred_rpart_PC)
cM_rpart_PC_valid <- confusionMatrix(as.factor(valid$classe), 
                                     predict(mdl_rpart_PC, validPC))
```


```{r models}


mdl_rpart <- train(classe ~ ., 
                   method = "rpart", 
                   trControl = trainControl(method = "cv", number = 3),
                   data = train_subset_2)
mdl_rf <- randomForest(as.factor(classe) ~ ., data = train_subset_2)

pred_rpart <- predict(mdl_rpart, train_subset_2)
pred_rf <- predict(mdl_rf, train_subset_2)
cM_rpart <- confusionMatrix(as.factor(train_subset_2$classe), pred_rpart)
cM_rf <- confusionMatrix(as.factor(train_subset_2$classe), pred_rf)

cM_rpart_valid <- confusionMatrix(as.factor(valid$classe), 
                                  predict(mdl_rpart, valid))
cM_rf_valid <- confusionMatrix(as.factor(valid$classe), 
                                  predict(mdl_rf, valid))
model_accuracy <- data.frame(
  cbind("Model" = c("rpart with PCA", "rpart", "Random Forest"),
        rbind(cM_rpart_PC$overall[1],
              cM_rpart$overall[1],
              cM_rf$overall[1]))) %>%
  mutate(Accuracy = as.numeric(Accuracy))
gt(model_accuracy, caption = "Model in-sample accuracy") %>%
  fmt_number(decimals = 3)
```

```{r}
OOS <- data.frame(
  cbind("Model" = c("rpart with PCA", "rpart", "Random Forest"),
        rbind(1-cM_rpart_PC_valid$overall[1],
              1-cM_rpart_valid$overall[1],
              1-cM_rf_valid$overall[1]))) %>%
  mutate("Error Rate" = as.numeric(Accuracy), .keep = "unused")

gt(OOS, 
          caption = "Out of sample Error Rate (1 - Accuracy)") %>%
  fmt_number(decimals = 3)
```



